---
title: "Project 11"
format: html
editor: visual
execute: 
  echo: false
  warning: false
---

```{r}
#| message: false
#| warning: false

library(tidyverse)    
library(ggplot2)      
library(performance)
library(skimr)        
library(sjPlot)
library(kableExtra)  
library(patchwork)
```

## Introduction

In this statistical report we analyse the trends in obesity prevalence based on socio-ecomic and lifestyle factors from 2013-2016 Scottish Health surveys. This report will also summarise key statistics and use plots to visualise the relationship between obesity rates, age, sex, and employment status.

## Exploratory Data Analysis

Before conducting formal statistical modelling, an exploratory analysis was performed to understand the distribution of obesity within the data-set and to identify potential patterns across demographic groups.

```{r}
data <- read.csv("DAproject11.csv")
```

#### Creating Obesity Indicator Varibale

Firstly, we create a binary response variable for obesity classification from the *BMIgroup* variable. This is then used to produce the plots and tables below.

```{r}
#|echo: true
data$Obese <- ifelse(data$BMIgroup == "Obese",1,0)

data$ObeseRate <- ifelse(data$BMIgroup == "Obese","yes","no")
```

#### Obesity Prevelance by Year

To examine whether obesity prevalence has changed between 2013 and 2016, the proportion of individuals classified as obese was calculated for each survey year.

```{r}
#|echo: true
mean(data$Obese)
obesity_year <- data %>%
  group_by(Year) %>%
  summarise(
    prevalence = mean(Obese)
  )

```

```{r}
ggplot(obesity_year, aes(x = Year, y = prevalence)) +
  geom_point() +
  geom_line()+ labs(title = " Figure___: Obesity Prevalance in Scotland by year", x = "Year", y = "Obesity Prevalence")
```

Figure \_\_ shows whether obesity prevalence has increased, decreased or remained stable over the survey years. To examine whether obesity prevalence has changed between 2013 and 2016, the proportion of individuals classified as obese was calculated for each survey year.

#### Obesity by Sex, Age and Employment

```{r}
#| fig-width: 12
#| fig-height: 8

library(tidyverse)

long_data <- data %>%
  pivot_longer(cols = c(AgeGroup, Sex, Employment),
               names_to = "Variable",
               values_to = "Category")

obesity_rates <- long_data %>%
  group_by(Variable, Category) %>%
  summarise(ObesityRate = mean(Obese), .groups = "drop")

q1 <- ggplot(obesity_rates,
       aes(x = factor(Category, levels = c("16-24","25-34","35-44","45-54","55-64","65-74","75+",
   "Male","Female",
   "Full-time education",
   "Paid employment",
   "Looking for work",
   "Looking after home/family",
   "Retired",
   "Permanently unable to work")), y = ObesityRate, colour = Variable, group = Variable)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Figure : ___Obesity Prevalence by Age Group, Sex and Employment",
       x = "Category",
       y = "Obesity Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




long_data$Variable <- factor(long_data$Variable,
                             levels = c("AgeGroup","Sex","Employment"))

long_data$Category <- factor(long_data$Category,
 levels = c(
   "16-24","25-34","35-44","45-54","55-64","65-74","75+",
   "Male","Female",
   "Full-time education",
   "Paid employment",
   "Looking for work",
   "Looking after home/family",
   "Retired",
   "Permanently unable to work"
 ))

q2 <- ggplot(long_data, aes(x = Category, y = Obese, fill = Variable)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs( title = "Figure ___ : Obesity Prevalence by Sex, Age and Employment status")  

(q1|q2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

Figure\_\_: shows ...

Figure \_\_ indicates that age is associated with differences in obesity prevalence. Obesity rates appear to increase progressively across the younger and middle age groups, reaching their highest levels in the *65–74* category, before declining slightly in the *75+* group.

The plot also suggests small differences between sexes, with females exhibiting slightly higher obesity rates on average than males.

Finally, the figure indicates that employment status is related to obesity prevalence. Individuals in the categories “*Looking after home/family*” and “*Retired*” show comparatively higher obesity rates than those in other employment categories.

```{r}
#| fig-width: 15
#| fig-height: 8

p1<- ggplot(data, aes(x = AgeGroup, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Prevalence by Age Group and Sex",
       x = "Age Group",
       y = "Obesity Rate") +
  theme_minimal()

p2 <- ggplot(data, aes(x = AgeGroup, y = Obese, fill = Employment)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Rate by Age Group and Employment",
       x = "Age Group",
       y = "Obesity Rate") +
  theme_minimal()  + guides(x = guide_axis(angle = 90))




(p1| p2 ) + 
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

```{r}
p3 <- ggplot(data, aes(x = Employment, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Prevalence by Employment and Sex",
       x = "Employment Status",
       y = "Obesity Prevalence") +
  theme_minimal()  + guides(x = guide_axis(angle = 90))

p3
```

Figure \_\_ shows that obesity prevalence differ slightly between males and females across age groups. In most age groups up to and including *45–54*, as well as in the *75+* category, females exhibit higher obesity rates than males. In contrast, in the *55–64* age group, males have a higher obesity rate than females. For individuals aged *65–74*, the obesity rates for males and females are very similar, with little difference between the two groups.

Figure\_\_ shows POSSIBLY DO NOT INCLUDE - TOO BUSY?

Figure\_\_ show that obesity rates vary across categories and between sexes. In most emploment groups, females have higher obesity rates than males. An exception occurs in *Looking after home/family* where males have a higher obesity rate.

Additionally, obesity rates among individuals who are retired are very similar for males and females. This is consistent with the findings in figure \_\_ (one 2 above), where obesity rates for individuals aged *65-74* were also observed to be very similar between males and females.

#### Obesity by Sex, Age Group or Employment Status

```{r}
#| fig-width: 15
#| fig-height: 8

r1 <- ggplot(data, aes(x = Sex, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs(
    title = "Figure __: Obesity prevalence by sex",
    y = "Proportion",
    fill = "ObeseRate"
  )

r2 <- 
ggplot(data, aes(x = AgeGroup, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs(
    title = "FIgure __: Obesity prevalence by age group",
    y = "Proportion",
    fill = "ObeseRate"
  )

r3 <-  ggplot(data, aes(x = Employment, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs(
    title = "Figure___: Obesity prevalence by employment status",
    y = "Proportion",
    fill = "ObeseRate"
  ) + guides(x = guide_axis(angle = 90))

(r1|r2|r3) +
   plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

The distribution of obesity across males and females was also examined. The visualisation findings in figure\_\_\_ show that obesity prevalence differs slightly between sexes, although the difference appears relatively small. Further statistical analysis is required to determine whether this difference is statistically significant.

Figure \_\_ demonstrates substantial variation in obesity prevalence can be observed across age groups. Older age categories tend to show a higher proportion of individuals classified as obese compared with younger age groups. This suggests that age may be an important factor associated with obesity risk.

Figure \_\_ shows that the differences in obesity prevalence are also visible across employment categories. These differences may reflect underlying Social and economic factors that influence lifestyle behaviours and health outcomes.

#### Obesity Prevalence by Sex, Age Group and Employment Status

```{r}
#|echo: false 
#|warning: false
s1 <- data%>%
  group_by(Sex)%>%
  summarise(prevalence = round(mean(Obese),3)) %>%
  kable(col.names = c("Sex", "Obesity Prevalence"),caption = "Table 1: Obesity Prevalence by Sex")

s2 <- data %>%
  group_by(AgeGroup) %>%
  summarise(prevalence = round(mean(Obese),3))%>%
  kable(col.names = c("Age Group", "Obesity Prevalence"), caption = "Table 2: Obesity Prevalence by Age Group")

s3 <- data %>%
  group_by(Employment) %>%
  summarise(obesity_rate = round(mean(Obese),3))%>%
  kable(col.names = c("Employment Status", "Obesity Prevalence"), caption = "Table 3: Obesity Prevalence by Employment Status")
```

::: columns
::: {.column width="25%"}
```{r}
s1
```
:::

::: {.column width="25%"}
```{r}
s2
```
:::

::: {.column width = "50%"}

```{r}
s3
```
:::

:::

Table 1 shows that obesity prevalence is slightly higher among females, with an obesity rate of 0.351, compared with 0.291 among males.

Table 2 also indicates that the *16–24* age group has the lowest obesity rate. Obesity prevalence increases progressively with age, reaching a peak in the *65–74* age group, before declining slightly in the *75+* age group.

Table 3 further shows variation by employment status. Individuals in *full-time employmen*t have the lowest obesity rate (0.160), while those *permanently unable to work* exhibit the highest rate (0.426).

#### Table with all

```{r}
library(dplyr)
library(knitr)

table_data <- bind_rows(
  data %>%
    group_by(Sex) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Sex", Category = Sex),

  data %>%
    group_by(AgeGroup) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Age Group", Category = AgeGroup),

  data %>%
    group_by(Employment) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Employment Status", Category = Employment)
) %>%
  select(Variable, Category, prevalence)

kable(
  table_data,
  digits = 3,
  col.names = c("Variable", "Category", "Obesity Prevalence"),
  caption = "Table 4: Obesity prevalence by sex, age group, and employment status"
)
```

Table 4 shows that obesity prevalence is slightly higher among females, with an obesity rate of 0.351, compared with 0.291 among males.

Table 4 also indicates that the *16–24* age group has the lowest obesity rate. Obesity prevalence increases progressively with age, reaching a peak in the *65–74* age group, before declining slightly in the *75+* age group.

Table 4 further shows variation by employment status. Individuals in *full-time employmen*t have the lowest obesity rate (0.160), while those *permanently unable to work* exhibit the highest rate (0.426).

#### BMI Distribution by Demographic and Socio-Economic factors

```{r}
library(tidyverse)

plot_data <- data %>%
  pivot_longer(cols = c(Sex, AgeGroup, Employment),
               names_to = "variable",
               values_to = "category")

ggplot(plot_data, aes(x = category, fill = BMIgroup)) +
  geom_bar(position = "fill") +
  facet_wrap(~variable, scales = "free_x") +
  labs(title = "Figure__ : BMI Distribution by Demographic and Socio-economic Factors",
       x = "",
       y = "Proportion",
       fill = "BMI Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure\_\_ shows the proportion of individuals who fall into each BMI group against Age, Employment status and Sex. It indicates that indivduals in the !In full-time employment" category have the lowest obesity rate, whilst those in ""Permanently unable to work exhibit the highest obesity rate.

Figure \_\_ also shows that the second oldest age group (65-74) have the highest rate of obesity, whilst the youngest age group (16-24) has the lowest rate. the plot also demonstrates that females have a slightly higher obesity rate than males.

#### BMI vs

BMI vs Sex

```{r}
table(data$BMIgroup, data$Sex)
```

BMI vs Age

```{r}
table(data$BMIgroup, data$AgeGroup)
```

BMI vs Employment

```{r}
table(data$BMIgroup, data$Employment)
```

#### Chi- Square Tests (NOT EDA??)

The Chi-Square test of independence was conducted to examine whether BMI group is associated with Sex, Age and Employment status

```{r}
chisq.test(table(data$BMIgroup, data$Sex))

```

The results from this chi-square test show a chi-square statistic of 190.19 with 3 degrees of freedom, and a p-value of $<2.2 \times10^{-16}$ . This is an extreemly small p-value and is well below the conventional signifance level of 0.05, indicating string evidence agsinst the null hypothesis of independence.

```{r}
chisq.test(table(data$BMIgroup, data$AgeGroup))


```

The results from this chi-square test show a chi-square statistic of 743.05 with 18degrees of freedom, and a p-value of $<2.2 \times10^{-16}$. This is an extremely small p-value and is well below the conventional signiificance level of 0.05, indicating string evidence against the null hypothesis of independence.

```{r}
#| warnings: false
chisq.test(table(data$BMIgroup, data$Employment))
```

The results from this chi-square test show a chi-square statistic of 506.42 with 18 degrees of freedom, and a p-value of $<2.2\times 10^{-16}$. This is an extremely small p-value and is well below the conventional significance level of 0.05, indicating string evidence againstm the null hypothesis of independence.

Overall, the exploratory analysis highlights potential relationships between obesity prevalence and several demographic characteristics. These patterns will be formally tested in the following section using statistical modelling.
