---
title: "Obesity Prevalence in Scotland (2013–2016)"
author: "Group 11"
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false
#| warning: false

library(tidyverse)    
library(ggplot2)      
library(performance)
library(skimr)        
library(sjPlot)
library(kableExtra)   
```

# Introduction {#sec-intro}

Obesity is a major public health issue and has been linked to a range of chronic health conditions, including cardiovascular disease and diabetes. Monitoring obesity prevalence and identifying population groups at higher risk are therefore important for informing public health strategies and interventions.

This study analyses data from the Scottish Health Survey covering the period from 2013 to 2016. The dataset provides demographic and socio-economic information that allows an investigation of patterns in obesity prevalence across different segments of the population.

The main objective of this analysis is to examine whether the prevalence of obesity in Scotland changed during this period and to explore whether obesity rates differ across demographic characteristics such as age, sex, and employment status. To address these questions, we first conduct an exploratory analysis to identify potential patterns in the data. This is followed by a formal statistical analysis using a logistic regression model to assess the relationship between obesity and the explanatory variables.

```{r}
library(tidyverse)

data <- read.csv("DAProject11.csv")
```

# Exploratory Analysis {#sec-EA}

Before conducting formal statistical modelling, an exploratory analysis was performed to understand the distribution of obesity within the data-set and to identify potential patterns across demographic groups.

```{r}
data <- na.omit(read_csv("C:/Users/joelt/Documents/Course Content/Stats 3/Data Analysis/Group Project/DAProject11.csv"))

data$obesity <- (ifelse(data$BMIgroup == "Obese",1,0))

```

```{r}
#| echo: false
#| message: false

obesity_year <- data %>% group_by(Year) %>%
  summarise(Rate = round(mean(as.numeric(obesity)),3), Count = n())

obesity_age <- data %>% group_by(AgeGroup) %>%
  summarise(Rate = round(mean(as.numeric(obesity)),3), Count = n())

obesity_emp <- data %>% group_by(Employment) %>%
  summarise(Rate = round(mean(as.numeric(obesity)),3), Count = n())

obesity_sex <- data %>% group_by(Sex) %>%
  summarise(Rate = round(mean(as.numeric(obesity)),3), Count = n())

mean(data$obesity)
sd(data$obesity)
```

The overall rate of obesity across all measured data was 0.304 with a standard deviation of 0.460.

```{r}
#| echo: false
#| label: tbl-obesity_year
#| fig-align: center
#| fig-cap: Table of Obesity Prevalence in Each Year
obesity_year %>% 
  gt(rowname_col = "Year") %>%
  tab_header(
    title = "Obesity Prevalence by Year"
  ) %>%
  tab_stubhead(label="Year")

```

```{r}
#| echo: false
#| label: fig-obesity_year_bar
#| fig-align: center

ggplot(data, aes(fill=as.factor(obesity),x=Year)) +
  geom_bar() +
  labs(title="Obesity by Year", y="Count", fill="Obesity")


```

We begin by considering the changes in obesity prevalence across the years in question. It appears that from @fig-obesity_year_bar and @tbl-obesity_year that there have been minimal variations in rate across the years, and since the sample size is split fairly evenly between years, these results are likely evenly representative. Prevalence appears to have peaked in 2015, having been at its lowest point the prior year.

```{r}
#| echo: false
#| label: tbl-obesity_sex
#| fig-align: center
obesity_sex %>% 
  gt(rowname_col = "Sex") %>%
  tab_header(
    title = "Obesity Prevalence by Sex"
  ) %>%
  tab_stubhead(label="Sex")

```

```{r}
#| echo: false
#| label: fig-obesity_sex_bar
#| fig-align: center
ggplot(data, aes(fill=as.factor(obesity),x=Sex)) +
  geom_bar() +
  labs(title="Obesity by Sex", y="Count", fill="Obesity")


```

We can see from @tbl-obesity_sex that the rate of obesity appears to be slightly higher in female participants than male.

```{r}
#| echo: false
#| label: tbl-obesity_age
#| fig-align: center
obesity_age %>% 
  gt(rowname_col = "AgeGroup") %>%
  tab_header(
    title = "Obesity Prevalence by Age"
  ) %>%
  tab_stubhead(label="Age")
  
```

```{r}
#| echo: false
#| label: fig-obesity_age_bar
#| fig-align: center
ggplot(data, aes(fill=as.factor(obesity),x=AgeGroup)) +
  geom_bar() +
  labs(title="Obesity by Age", y="Count", fill="Obesity")


```

We can see an increase in obesity rate as we ascend age groups, with all but 75+ being higher than the last. Obesity rates appear to peak between ages 65-74, however the rates are similar between 45-54, 55-64 and 65-74. This lines up with what one would expect from age, as physical ability and activity tend to decrease with age, as well as higher prevalence of autoimmune disease and hormonal conditions. The drop in prevalence among 75+ year-olds is also expected as elderly people's diets are often nutritionally insufficient.

```{r}
#| echo: false
#| label: tbl-obesity_emp
#| fig-align: center
obesity_emp %>% 
  gt(rowname_col = "Employment") %>%
  tab_header(
    title = "Obesity Prevalence by Employment Status"
  ) %>%
  tab_stubhead(label="Status")

```

```{r, fig.width=10}
#| echo: false
#| label: fig-obesity_emp_bar
#| fig-align: center

ggplot(data, aes(fill=as.factor(obesity),y=fct_rev(Employment))) +
  geom_bar() +
  labs(title="Obesity by Employment Status", y="Count", fill="Obesity")
  


```

Obesity rates vary greatly with employment status, though so does share of the study in each group. There is comparatively little data for groups like those "Doing something else" while over half of all data pertains to those "In paid employment etc.". The highest rates appear to be among those permanently unable to work, which aligns with expectation as the reason for this tends to be disability which can have significant correlations with weight extremes. The lowest rates are among those in full time education, which is in line with expectations also as most students are in younger age groups which we know from @tbl-obesity_age have the lowest obesity rates.

# Formal Analysis {#sec-FA}

To formally examine the determinants of obesity, we estimate a **logistic regression model** where the probability of being obese depends on demographic characteristics and survey year.

```{r}
#|echo: false
model <- glm(
  Obese ~ Year + Sex + AgeGroup + Employment,
  data = data,
  family = binomial
)

tab_model(model)
```

The logistic regression model allows us to estimate how different factors influence the likelihood of obesity.

Key findings include:

-    The coefficient for **Year** indicates whether obesity prevalence has increased or decreased over time.

-   Differences between **Sex** categories indicate whether obesity risk varies between males and females.

-    **Age-Group** coefficients capture differences across age ranges.

-   **Employment status** may reflect social economic influences on obesity risk.

-    Odds ratios provide an interpretation in terms of relative risk.

```{r}
exp(coef(model))
```

These values represent the multiplicative change in the odds of obesity associated with each predictor.

```{r}
chisq.test(table(data$Sex,data$Obese))
```

# Conclusions {#sec-con}

This analysis investigated obesity prevalence in Scotland using data from the Scottish Health Survey (2013–2016).

The exploratory analysis showed variation in obesity prevalence across demographic groups, particularly across different age groups and employment categories.

The logistic regression model confirmed that demographic characteristics are associated with obesity risk. These findings suggest that obesity is not evenly distributed across the population and may be influenced by social and economic and demographic factors.

Future work could extend this analysis by incorporating additional lifestyle variables such as diet, physical activity, or income levels. Analysing more recent survey data would also help determine whether the observed patterns persist over time.
