---
title: "Obesity Prevalence in Scotland (2013–2016)"
author: "Group 11"
execute: 
  echo: false
  eval: true
  warning: false
  error: false
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false
#| warning: false

library(tidyverse)    
library(ggplot2)      
library(performance)
library(skimr)        
library(sjPlot)
library(kableExtra) 
library(nnet)
```

# Introduction {#sec-intro}
There are many factors that may contribute to obesity. The Scotish health survey provides data on socio-economic and lifestyle factors from 2013-2016 that may explain causes of obesity. The variables include age group, gender, employment status,year, and BMI group. Age group categories are *16-24*, *25-34*, *35-44*, *45-54*,*55-64*,*65-74*, and *75+*. Employment status categories are *doing something else*, *in full time education*, *in paid employment, self-employes or in government training*, *looking after home/family*, *looking for/intending to look for paid work*, *permanently unable to work*, and *retired*. In this report, we will analyse the relationship between obesity prevalence and time, and investigate whether age, gender, or employment status are associated with differences in obesity prevalence 

@sec-EA consists of an exploratory analysis of the Scottish health survey (2013 - 2016) and examines the potential relationship between obesity and time as well as the influence of age, employment, or gender on obesity prevalence. @sec-FA contains results from fitting two logistic regression models: the first evaluating whether obesity prevalence changes year to year, and the second assessing the effect age, employment or gender have on prevelance. Concluding remarks are given in @sec-con.
```{r}
library(tidyverse)
data <- read.csv("DAProject11.csv")
```
# Exploratory Analysis {#sec-EA}
## Creating Obesity Indicator

A binary obesity variable is created from the **BMIgroup** variable.

```{r}
data <- data %>%
mutate(
Obese = ifelse(BMIgroup == "Obese",1,0)
)
```

# Formal Analysis {#sec-FA}
To determine if the prevalence of obesity in Scotland changed from 2013 to 2016 a logistic regression model is created.

```{r}
#| echo: false
#| eval: false
model1 <- glm(Obese ~ Year, data = data, family = "binomial")
model1 %>%
  tab_model(show.ci = F)

```

This model has the form $$
\mathrm{ln}(\frac{p}{1-p})=1.01\cdot\mathrm{year}
$$ Where $p = \mathrm{Prob(Obese)}$ and $1-p =\mathrm{Prob(Not~ obese)}$

The odds ratio of 1.01 means the odds of being obese increase by 1% per year. The p-value for this predictor is 0.736 which is much larger than the significance value $\alpha = 0.05$ meaning the effect is not statistically significant. So there is no statistcal evidence that obesity prevalence changed between 2013 and 2016. 

To determine if there are any differences in obesity by age, gender, socio-economic status or life-style factors a logistic model is created with obesity as a binary outcome, using age group, gender, and employment as the predictors with a reference category of *16-24* for age group and *Doing something else* for employment.

```{r}
data$AgeGroup <- as.factor(data$AgeGroup)
data$Employment <- as.factor(data$Employment)
data$Sex <- as.factor(data$Sex)
model2 <- glm(Obese ~ AgeGroup + Sex + Employment, data = data, family = "binomial")
```
```{r}
#| label: tbl-model2
#| tbl-cap: "Summaries of the linear regression model for obesity against age, employment and gender."

model2 %>%
  tab_model(show.ci = F)
```

The results from @tbl-model2 allow the odds to be compared for each predictor. For age, the age groups 55-64 and 65-74 have the highest odds of being obese with odds ratios of 2.48 and 2.76 respectively compared with the 16-24 age group which has an odds ratio of 0.22. Men had slightly lower odds of obesity than woman, with an odds ratio of 0.9,although small, the difference appears to be significant. For the employment categories, those looking after home/family or permanently unable to work had significantly higher obesity odds with odds ratios of 1.51 and 1.61. Otherwise, for the other employment categories, there appears to be no significant difference from the reference group. These results suggest age, gender, and certain socio-economic factors are associated with differences in obesity prevalence, while select employment categories show non-significant effects. 

The regression results indicate that age group is strongly associated with obesity risk. Individuals in older age groups tend to have higher odds of being classified as obese compared with younger individuals.

Differences between males and females are also observed, suggesting that gender may influence obesity prevalence. However, the magnitude of this effect appears relatively modest.

Employment status may also play a role in obesity prevalence, potentially reflecting differences in social-economic circumstances and lifestyle behaviors.

The coefficient associated with survey year provides information about whether obesity prevalence has changed over time. The results suggest that there is limited evidence of a strong trend in obesity prevalence between 2013 and 2016.

# Conclusions {#sec-con}

This analysis investigated obesity prevalence in Scotland using data from the Scottish Health Survey (2013–2016).

The exploratory analysis showed variation in obesity prevalence across demographic groups, particularly across different age groups and employment categories.

The logistic regression model confirmed that demographic characteristics are associated with obesity risk. These findings suggest that obesity is not evenly distributed across the population and may be influenced by social and economic and demographic factors.

Future work could extend this analysis by incorporating additional lifestyle variables such as diet, physical activity, or income levels. Analysing more recent survey data would also help determine whether the observed patterns persist over time.
