---
title: "Obesity Prevalence in Scotland (2013–2016)"
author: "Group 11"
execute: 
  echo: false
  eval: true
  warning: false
  error: false
number-sections: true
format:
  pdf:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false
#| warning: false

library(tidyverse)    
library(ggplot2)      
library(performance)
library(skimr)        
library(sjPlot)
library(kableExtra) 
library(nnet)
library(see)
library(patchwork)
```

# Introduction {#sec-intro}
There are many factors that may contribute to obesity. The Scottish health survey provides data on socio-economic and lifestyle factors from 2013-2016 that may be associated with obesity. The variables include age group, gender, employment status,year, and BMI group. Age group categories are *16-24*, *25-34*, *35-44*, *45-54*,*55-64*,*65-74*, and *75+*. Employment status categories are *Doing something else*, *In full time education*, *In paid employment, self-employed or in government training*, *Looking after home/family*, *Looking for/intending to look for paid work*, *Permanently unable to work*, and *Retired*.  In this report, we will analyse the relationship between obesity prevalence and surveyed years. In addition we will investigate whether age, gender, or employment status are associated with differences in obesity prevalence. 

@sec-EA consists of an exploratory analysis of the Scottish health survey (2013 - 2016) and examines the potential relationship between obesity and time as well as the influence of age, employment, or gender on obesity prevalence. @sec-FA contains results from fitting two logistic regression models: the first evaluating whether obesity prevalence changes year to year, and the second assessing the effect age, employment, or gender have on prevalence. Concluding remarks are given in @sec-con.
```{r}
data <- read.csv("DAProject11.csv")
```
# Exploratory Analysis {#sec-EA}

```{r}
data <- read.csv("DAproject11.csv")
```

```{r}
#|echo: true
data$Obese <- ifelse(data$BMIgroup == "Obese",1,0)

data$ObeseRate <- ifelse(data$BMIgroup == "Obese","yes","no")
```

#### Obesity Prevalence in Scotland by Year

To examine whether obesity prevalence has changed between 2013 and 2016, the proportion of individuals classified as obese was calculated for each survey year.

```{r}
#|echo: false
#|output: false
mean(data$Obese)
obesity_year <- data %>%
  group_by(Year) %>%
  summarise(
    prevalence = round(mean(Obese),3)
  )
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| label: tbl-1
#| tbl-cap: "Percentage of obese individuals surveyed per given year, between 2013 and 2016." 
year_rates <- data %>%
  group_by(Year) %>%
  summarise(ObesityRate = round(mean(Obese)*100,1))
colnames(year_rates) <- c("Year", "Obesity Rates")
kable(year_rates)
```

@tbl-1 shows the obesity rates by year in Scotland between 2013 and 2016. The percentage of individuals classified as obese fluctuated between 30.2% and 30.6%, indicating no clear upward or downward trend over the surveyed years.

```{r}
#| fig-width: 6
#| fig-height: 4
#| label: fig-1
#| fig-cap: "Graph of obesity prevelance in Scotland by year"
ggplot(obesity_year, aes(x = Year, y = prevalence)) +
  geom_point() +
  geom_line()+ labs(title = " Obesity Prevalance in Scotland by year", x = "Year", y = "Obesity Prevalence")
```

@fig-1 shows how obesity prevalence changes over the given years. To examine this the proportion of individuals classified as obese is calculated for each survey year. Similary to the results found in @tbl-1, there is no significant upwards or downwards trend over the surveyed years.


```{r}
#| fig-width: 15
#| fig-height: 8
#| label: fig-2
#| fig-cap: "Obesity Prevalence by Demographic Variables"


p1<- ggplot(data, aes(x = AgeGroup, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title= "Obesity Prevalence by Age Group and Sex",  x = "Age Group",
       y = "Obesity Rate") +
  theme_minimal()

p2 <- ggplot(data, aes(x = Employment, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Obesity prevalence by Employment and Sex",
       x = "Employment Status",
       y = "Obesity Prevalence") +
  theme_minimal()  + guides(x = guide_axis(angle = 40))




(p1| p2 ) + 
  plot_annotation(
    tag_levels = list(c("(a)",
                   "(b)"))) &
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

@fig-2 indicates that age is associated with differences in obesity prevalence. Disregarding sex obesity rates appear to increase progressively across the younger and middle age groups, reaching their highest levels in the *65–74* category, before declining slightly in the *75+* group.
@fig-2 (panel a) shows that obesity prevalence differ slightly between males and females across age groups. In most age groups up to and including *45–54*, as well as in the *75+* category, females exhibit higher obesity rates than males. In contrast, in the *55–64* age group, males have a higher obesity rate than females. For individuals aged *65–74*, the obesity rates for males and females are very similar, with little difference between the two groups.

@fig-2 (panel b) show that obesity rates vary across categories and between sexes. In most employment groups, females have higher obesity rates than males. An exception occurs in *Looking after home/family* where males have a higher obesity rate. Additionally, obesity rates among individuals who are retired are very similar for males and females. This is consistent with the findings in @fig-2 (panel a), where obesity rates for individuals aged *65-74* were also observed to be very similar between males and females.

#### Obesity by Sex, Age Group or Employment Status

```{r}
#| fig-width: 15
#| fig-height: 8
#| label: fig-3
#| fig-cap: "Graph of Obesity prevalence by Demographic varibales"

r1 <- ggplot(data, aes(x = Sex, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs( title = "Obesity Prevalence by Sex",
    y = "Proportion",
    fill = "ObeseRate"
  )

r2 <- 
ggplot(data, aes(x = AgeGroup, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs( title = "Obesity Prevalence by Age group",
    y = "Proportion",
    fill = "ObeseRate"
  )

r3 <-  ggplot(data, aes(x = Employment, fill = factor(ObeseRate))) +
  geom_bar(position = "fill") +
  labs( title = "Obesity Prevalence by Employment Status",
    y = "Proportion",
    fill = "ObeseRate"
  ) + guides(x = guide_axis(angle = 40))

(r1|r2|r3) +
  plot_annotation(tag_levels = list(c("(a)",
                   "(b)", "(c)"))) &
   plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
```

The distribution of obesity across males and females was also examined. The visualisation findings in @fig-3 (panel a) show that obesity prevalence differs slightly between sexes, although the difference appears relatively small. Further statistical analysis is required to determine whether this difference is statistically significant.

@fig-3 (panel b) demonstrates substantial variation in obesity prevalence can be observed across age groups. Older age categories tend to show a higher proportion of individuals classified as obese compared with younger age groups. This suggests that age may be an important factor associated with obesity risk.

@fig-3 (panel c) shows that the differences in obesity prevalence are also visible across employment categories. These differences may reflect underlying Social and economic factors that influence lifestyle behaviours and health outcomes.

#### Obesity Prevalence by Sex, Age and Employment Status

```{r}
#| label: tbl-2
#| tbl-cap: "Obesity prevalence by Sex, Age Group, and Employment Status"
library(dplyr)
library(knitr)

table_data <- bind_rows(
  data %>%
    group_by(Sex) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Sex", Category = Sex),

  data %>%
    group_by(AgeGroup) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Age Group", Category = AgeGroup),

  data %>%
    group_by(Employment) %>%
    summarise(prevalence = mean(Obese)) %>%
    mutate(Variable = "Employment Status", Category = Employment)
) %>%
  select(Variable, Category, prevalence)

kable(
  table_data,
  digits = 3,
  col.names = c("Variable", "Category", "Obesity Prevalence")
)
```

@tbl-2 shows that obesity prevalence is slightly higher among females, with an obesity rate of 0.351, compared with 0.291 among males.

@tbl-2 also indicates that the *16–24* age group has the lowest obesity rate. Obesity prevalence increases progressively with age, reaching a peak in the *65–74* age group, before declining slightly in the *75+* age group.

@tbl-2 further shows variation by employment status. Individuals in *full-time employmen*t have the lowest obesity rate (0.160), while those *permanently unable to work* exhibit the highest rate (0.426).

Overall, the exploratory analysis highlights potential relationships between obesity prevalence and several demographic characteristics. These patterns will be formally tested in the following section using statistical modelling.


# Formal Analysis {#sec-FA}
To determine if the prevalence of obesity in Scotland changed from 2013 to 2016 a logistic regression model is created. 

```{r}
model1 <- glm(factor(Obese) ~ factor(Year), data = data, family = "binomial")
```
```{r}
#| eval: TRUE
model1 %>%
  tab_model(show.ci = F)
```
The odds ratios of 0.99, 1.02, and 1.01 indicates the odds of being obese are estimated to increase by about 1% per year. Which is quite minimal. Additionally, the p-values of 0.913, 0.749, and 0.832 are much larger than the significance value $\alpha = 0.05$ meaning the effect is not statistically significant. So, there is not enough statistical evidence to believe obesity prevalence changed between 2013 and 2016 based on the sample. 

To assess model fit and check assumptions, the following plots are made:
```{r}
#| label: fig-assump1
#| fig-cap: "Plots for checking the assumptions of the model."
#| fig-width: 10
#| fig-height: 8
check_model(model1, check = c("pp_check","binned_residuals","outliers","qq"))
```
The plots in @fig-assump1 check the assumptions for the logistic regression model. The posterior predictive plot (top left) indicates the model is able to reproduce the observed data, suggesting a good fit overall. The binned residuals plot (top right) shows the residuals fall within the error bounds, implying the link function is suitable. The influential observations plot (bottom left) shows there are no highly influential data points, as they all lie within the contour lines. The QQ-plot of the quantile residuals (bottom right) shows that the residuals fall closely on the line showing no major deviations from the expected residual behaviour.

Since no major violations are evident, the logistic regression model appears to be a good fit for the data. 

To determine if there are any differences in obesity by age, gender, socio-economic status or life-style factors a logistic model is created with obesity as a binary outcome, using age group, gender, and employment as the predictors with a reference category of *16-24* for age group and *Doing something else* for employment.

```{r}
data$AgeGroup <- as.factor(data$AgeGroup)
data$Employment <- as.factor(data$Employment)
data$Sex <- as.factor(data$Sex)
model2 <- glm(Obese ~ AgeGroup + Sex + Employment, data = data, family = "binomial")
```
```{r}
#| label: tbl-model2
#| tbl-cap: "Summaries of the logistic regression model for obesity against age, employment and gender."

model2 %>%
  tab_model(show.ci = F)
```

The results from @tbl-model2 allow the odds to be compared for each predictor. The regression results indicate that age group is strongly associated with obesity risk, more specifically, the age groups 55-64 and 65-74 have the highest odds of being obese with odds ratios of 2.48 and 2.76 respectively compared with the 16-24 age group which has an odds ratio of 0.22. Differences between males and females are also observed, suggesting that gender may influence obesity prevalence where males have slightly lower odds of obesity compared to females, with an odds ratio of 0.9. Although small, the difference appears to be statistically significant, with a p-value less than 0.05. Employment status may also play a role in obesity prevalence, potentially reflecting differences in social-economic circumstances and lifestyle behaviors. Those looking after home/family or permanently unable to work had significantly higher obesity odds with odds ratios of 1.51 and 1.61. Otherwise, for the other employment categories, there appears to be no significant difference from the reference group. These results suggest age, gender, and some employment categories are associated with obesity, while select employment categories show non-significant effects. 

To determine if the model fits the data well, the following plots are made:
```{r}
#| label: fig-assump2
#| fig-cap: "Final model diagnostic plots."
#| fig-width: 10
#| fig-height: 8
check_model(model2, check = c("pp_check","binned_residuals","outliers","qq"))
```
The plots in @fig-assump2 assess the assumptions and fit of the logistic regression model. The posterior predictive check (top left) indicates the model captures the pattern of the observed data. The binned residuals plot (top right) shows the residuals mostly fall within the error bounds, suggesting an appropriate model specification. There are no indications of problematic observations as shown in the influential observations plot (bottom left), as all points lie within the contour lines. The QQ plot (bottom right) shows the residuals align closely with the reference line, indicating no substantial departures from the expected distribution.

Since no assumptions appear to be violated, the logistic regression model appears to be a good fit for the data. 

# Conclusions {#sec-con}
The prevalence of diabetes in Scotland has not significantly changed between 2013 and 2016. *Age*, *employment*, and *gender* are associated with obesity. Where age groups 55-64 and 65-74 have the highest odds of being obese, men are less likely to be obese than women, and those looking after home/family or permanently unable to work also have higher odds of being obese. 

