---
title: "Project 11"
format: html
editor: visual
execute: 
  echo: false
---

```{r}
#| message: false
#| warning: false

library(tidyverse)    
library(ggplot2)      
library(performance)
library(skimr)        
library(sjPlot)
library(kableExtra)  
library(patchwork)
```

## Introduction

In this statistical report we analyse the trends in obesity prevalence based on socio-ecomic and lifestyle factors from 2013-2016 Scottish Health surveys. This report will also summarise key statistics and use plots to visualise the relationship between obesity rates, age, sex, and employment status.

## Exploratory Data Analysis

Before conducting formal statistical modelling, an exploratory analysis was performed to understand the distribution of obesity within the data-set and to identify potential patterns across demographic groups.

```{r}
data <- read.csv("DAproject11.csv")
```

#### Creating Obesity Indicator Varibale 

Firstly, we create a binary response variable for obesity classification from the *BMIgroup* variable. This is then used to produce the plots below.

```{r}
#|echo: true
data$Obese <- ifelse(data$BMIgroup == "Obese",1,0)
```

#### Obesity Prevelance by Year

To examine whether obesity prevalence has changed between 2013 and 2016, the proportion of individuals classified as obese was calculated for each survey year.

```{r}
#|echo: true
mean(data$Obese)
obesity_year <- data %>%
  group_by(Year) %>%
  summarise(
    prevalence = mean(Obese)
  )

```

```{r}
ggplot(obesity_year, aes(x = Year, y = prevalence)) +
  geom_point() +
  geom_line()+ labs(title = " Figure___: Obesity Prevalance in Scotland by year", x = "Year", y = "Obesity Prevalence")
```

Figure \_\_ shows whether obesity prevalence has increased, decreased or remained stable over the survey years. To examine whether obesity prevalence has changed between 2013 and 2016, the proportion of individuals classified as obese was calculated for each survey year.

#### Obesity by Sex, Age and Employment

```{r}
library(tidyverse)

long_data <- data %>%
  pivot_longer(cols = c(AgeGroup, Sex, Employment),
               names_to = "Variable",
               values_to = "Category")

obesity_rates <- long_data %>%
  group_by(Variable, Category) %>%
  summarise(ObesityRate = mean(Obese), .groups = "drop")

ggplot(obesity_rates,
       aes(x = Category, y = ObesityRate, colour = Variable, group = Variable)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Figure : ___Obesity Prevalence by Age Group, Sex and Employment",
       x = "Category",
       y = "Obesity Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure\_\_: shows ...

```{r}
long_data$Variable <- factor(long_data$Variable,
                             levels = c("AgeGroup","Sex","Employment"))

long_data$Category <- factor(long_data$Category,
 levels = c(
   "16-24","25-34","35-44","45-54","55-64","65-74","75+",
   "Male","Female",
   "Full-time education",
   "Paid employment",
   "Looking for work",
   "Looking after home/family",
   "Retired",
   "Permanently unable to work"
 ))

ggplot(long_data, aes(x = Category, y = Obese, fill = Variable)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs( title = "Figure ___ : Obesity Prevalence by Sex, Age and Employment status")  
```

Figure \_\_ indicates that age is associated with differences in obesity prevalence. Obesity rates appear to increase progressively across the younger and middle age groups, reaching their highest levels in the *65–74* category, before declining slightly in the *75+* group.

The plot also suggests small differences between sexes, with females exhibiting slightly higher obesity rates on average than males.

Finally, the figure indicates that employment status is related to obesity prevalence. Individuals in the categories “*Looking after home/family*” and “*Retired*” show comparatively higher obesity rates than those in other employment categories.

```{r}
ggplot(data, aes(x = AgeGroup, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Prevalence by Age Group and Sex",
       x = "Age Group",
       y = "Obesity Rate") +
  theme_minimal()
```

Figure \_\_ shows that obesity prevalence differ slightly between males and females across age groups. In most age groups up to and including *45–54*, as well as in the *75+* category, females exhibit higher obesity rates than males. In contrast, in the *55–64* age group, males have a higher obesity rate than females. For individuals aged *65–74*, the obesity rates for males and females are very similar, with little difference between the two groups.

```{r}
ggplot(data, aes(x = AgeGroup, y = Obese, fill = Employment)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Rate by Age Group and Employment",
       x = "Age Group",
       y = "Obesity Rate") +
  theme_minimal()  + guides(x = guide_axis(angle = 90))
```

Figure\_\_ shows

```{r}
ggplot(data, aes(x = Employment, y = Obese, fill = Sex)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  labs(title = "Figure__: Obesity Prevalence by Employment and Sex",
       x = "Employment Status",
       y = "Obesity Prevalence") +
  theme_minimal()  + guides(x = guide_axis(angle = 90))
```

Figure\_\_ show that obesity rates vary across categories and between sexes. In most emploment groups, females have higher obesity rates than males. An exception occurs in *Looking after home/family* where males have a higher obesity rate.

Additionally, obesity rates among individuals who are retired are very similar for males and females. This is consistent with the findings in figure \_\_ (one 2 above), where obesity rates for individuals aged *65-74* were also observed to be very similar between males and females.

#### Obesity by Sex

```{r}
#|eval: false
#|echo: false
data%>%
  group_by(Sex)%>%
  summarise(prevalence = mean(Obese))
```

```{r}
ggplot(data, aes(x = Sex, fill = factor(Obese))) +
  geom_bar(position = "fill") +
  labs(
    title = "Obesity prevalence by sex",
    y = "Proportion",
    fill = "Obese"
  )

```

Table\_\_: Obesity Prevalence by Sex

| Sex    | Obesity Rate |
|--------|--------------|
| Female | 0.315        |
| Male   | 0.291        |

The distribution of obesity across males and females was also examined. The visualisation findings in figure\_\_\_ show that obesity prevalence differs slightly between sexes, although the difference appears relatively small. Further statistical analysis is required to determine whether this difference is statistically significant.

#### Obesity by Age Group

```{r}
#|echo: false
#|eval: false 
data %>%
  group_by(AgeGroup) %>%
  summarise(prevalence = mean(Obese))
```

```{r}


ggplot(data, aes(x = AgeGroup, fill = factor(Obese))) +
  geom_bar(position = "fill") +
  labs(
    title = "Obesity prevalence by age group",
    y = "Proportion",
    fill = "Obese"
  )
```

Table\_\_ : Obesity Prevalence by Age Group

| Age   | Obesity Rate |
|-------|--------------|
| 16-24 | 0.16         |
| 25-34 | 0.244        |
| 35-44 | 0.285        |
| 45-54 | 0.328        |
| 55-64 | 0.352        |
| 65-74 | 0.369        |
| 75+   | 0.292        |

Figure \_\_ demonstrates substantial variation in obesity prevalence can be observed across age groups. Older age categories tend to show a higher proportion of individuals classified as obese compared with younger age groups. This suggests that age may be an important factor associated with obesity risk.

#### Obesity by Employment status

```{r}
#|eval: false
#|echo: false 
data %>%
  group_by(Employment) %>%
  summarise(obesity_rate = mean(Obese))
```

```{r}

ggplot(data, aes(x = Employment, fill = factor(Obese))) +
  geom_bar(position = "fill") +
  labs(
    title = "Obesity prevalence by employment status",
    y = "Proportion",
    fill = "Obese"
  ) + guides(x = guide_axis(angle = 90))
```

Table\_\_: Obesity Prevalence by Employment Status

| Employment Status                                      | Obesity rate |
|--------------------------------------------------------|--------------|
| Doing something else                                   | 0.297        |
| In full-time education                                 | 0.16         |
| In paid employment, self-employed or on gov't training | 0.285        |
| Looing after home/family                               | 0.372        |
| Looking for/intending to look for paid work            | 0.320        |
| Permanently unable to work                             | 0.426        |
| Retired                                                | 0.340        |

Fugure \_\_ shows that the differences in obesity prevalence are also visible across employment categories. These differences may reflect underlying Social and economic factors that influence lifestyle behaviours and health outcomes.

```{r}
library(tidyverse)

plot_data <- data %>%
  pivot_longer(cols = c(Sex, AgeGroup, Employment),
               names_to = "variable",
               values_to = "category")

ggplot(plot_data, aes(x = category, fill = BMIgroup)) +
  geom_bar(position = "fill") +
  facet_wrap(~variable, scales = "free_x") +
  labs(title = "Figure__ : BMI Distribution by Demographic and Socio-economic Factors",
       x = "",
       y = "Proportion",
       fill = "BMI Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure\_\_ shows the porprotion of individuals who fall into each BMI group against Age, Employment status and Sex.

#### BMI vs

BMI vs Sex

```{r}
table(data$BMIgroup, data$ Sex)
```

BMI vs Age

```{r}
table(data$BMIgroup, data$AgeGroup)
```

BMI vs Employment

```{r}
table(data$BMIgroup, data$Employment)
```

#### Chi- Square Tests (optional)

The Chi-Square test of independence was conducted to examine whether BMI group is associated with Sex, Age and Employment status

```{r}
chisq.test(table(data$BMIgroup, data$Sex))

```

The results from this chi-square test show a chi-square statistic of 190.19 with 3 degrees of freedom, and a p-value of $<2.2 \times10^{-16}$ . This is an extreemly small p-value and is well below the conventional signifance level of 0.05, indicating string evidence agsinst the null hypothesis of independence.

```{r}
chisq.test(table(data$BMIgroup, data$AgeGroup))


```

The results from this chi-square test show a chi-square statistic of 743.05 with 18degrees of freedom, and a p-value of $<2.2 \times10^{-16}$. This is an extremely small p-value and is well below the conventional signiificance level of 0.05, indicating string evidence against the null hypothesis of independence.

```{r}
#| warnings: false
chisq.test(table(data$BMIgroup, data$Employment))
```

The results from this chi-square test show a chi-square statistic of 506.42 with 18 degrees of freedom, and a p-value of $<2.2\times 10^{-16}$. This is an extremely small p-value and is well below the conventional significance level of 0.05, indicating string evidence againstm the null hypothesis of independence.

Overall, the exploratory analysis highlights potential relationships between obesity prevalence and several demographic characteristics. These patterns will be formally tested in the following section using statistical modelling.
